---
title: "Relevant NBA Stats from 2020-2021 Season"
author: "Eduardo Salvador"
date: "May 4, 2022"
output:
  pdf_document:
    highlight: haddock
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    # toc: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
geometry: margin = 0.5in
editor_options:
  chunk_output_type: console
urlcolor: blue
---




```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE # Turn off warnings,
                      )




```

\clearpage

```{r include=FALSE}

library(tidyverse) 
library(pander)  # We used pander here, but some alternative libraries are below:
library(xtable)




```

# Introduction 

This project is to be used as a way to research each player's position to find a correlation between the data of the 2020-2021 NBA season to see how each position in the NBA has changed from the traditional playstyle. I wanted to be able to look at trends in the data that suggest which position has the most and least point conversion whether it be 3pt or 2pt and what position has on average the youngest and oldest players. As well as answering the questions as to which NBA team has scored the most amount of points on average and the least amount of points on average during this season. Be aware to not use this model for any type of sports betting since, all models are wrong, only some are useful. 

\vspace{.2in}
\hrule
\vspace{.2in}

# Description of the Data 

1. I got the data from Kaggle. The link I used was: https://www.kaggle.com/datasets/umutalpaydn/nba-20202021-season-player-stats
2. This data frame has all the NBA players from 2020 and 2021
3. Content:
  1.Per Game Stats: Basically dividing every individual stats by the played game .
  2.Per 36 Minute Stats: In order to calculate per-36 minute stats, you divide 36 by the number of minutes the player actually played, then take that number and multiply all of the player's stats by it.
  3.Advanced Stats: These are more focused on the players direct effect on winning the games or scoring a point. For example a key tenet for many modern basketball analysts is that basketball is best evaluated at the level of possessions. During a single game, both teams have approximately the same number of possessions, because they alternate possession. (A team can have slightly more if it begins and ends a quarter or half with possession.) However, over the course of the season, teams play at very different paces, which can dramatically color their points scored and points allowed per game. Therefore, these analysts favor use of points scored per 100 possessions (offensive rating) and points allowed per 100 possessions (defensive rating).
  
I choose to analyze the data from the NBA 2020-2021 season since I am a big basketball fanatic and this season was the first season I got into NBA fantasy which for those reading who don't know is getting together with friends to draft specific players and follow them through the whole season while playing every single week against one of your friend's teams and the one who has the best team wins. This also involves constantly looking at the roster since various players suffer from injuries as well as making sure the active players in your roster play that week. I became a big fan of this but, unfortunately, since it was my first time playing I did really poorly so, I want to analyze this dataset to find the key positions I should be focusing on drafting. 
  
  \vspace{.2in}
\hrule
\vspace{.2in}
  
# Methods


To answer the questions from the Introduction which are listed below the following statistical methods were used to solve them: 

1. What position has on average the youngest and oldest players?

For the first question, the statistical method used was multiple boxplots to showcase each position on the x-axis and age in years on the y-axis. This will show not only the mean age of each position but also the distribution of numerical data and skewness by displaying this data quartiles.

2. Which position has the most and least point conversion where it be 3pt or 2pt?

For the second question, the statistical method used was a combination between linear regression and scatterplot to see the correlation between the 3 pt and 2 pt attempts with the actual points scored between all the different positions. The scatterplot helps visualize the disparities between all the data points between all the different positions and the linear regression helps identify the most and least point conversion based on the slope of each line. 



3. Which NBA team has the most and least amount of points scored on average?

For the third question, the statistical method used was to create a histogram that has the Teams in the x value and the number of Points on average scored on the y value. With this, it will be easily identified which team has scored the most and least points on average for the 2020-2021 season. 



\vspace{.2in}
\hrule
\vspace{.2in}

# Results and Discussion

Multiple boxplot

In an initial investigation, I wanted to know the average player's age based on their position to see if there is a correlation between player position and the amount of experience. Based on the results, the graph shows that the center position in the NBA on average has the oldest players which makes sense since the center position's most important attribute is the height of a player and it is really difficult to find tall younger players. The graph also shows the youngest position is the point guard and the role of small forward and power forward players. This also makes sense since for this kind of position the most important attribute is athleticism and strength in which young players may have the advantage as compared to older players because age does affect this. 

```{r echo=FALSE}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(olsrr)
library(pander)

#Reading the file
dfadplayers<-read.csv("/Users/eduardosalvador/Desktop/FINAL\ Spring\ Semester\ 2021/CMDA\ /Assignments/Project\ 1/archive/nba2021_advanced.csv",header = T)

dfpergame<-read.csv("/Users/eduardosalvador/Desktop/FINAL\ Spring\ Semester\ 2021/CMDA\ /Assignments/Project\ 1/archive/nba2021_per_game.csv",header = T)

#Creating ggplot 
ggplot(dfadplayers,aes(x=Pos,y=Age,fill=Pos))+theme_bw()+geom_boxplot()+ggtitle("Average player age based on position")
mean(dfadplayers$Age)

 

 
```


Linear Regression + Scatterplot

I wanted to see if there was a correlation between 3 pt attempts, 2 pt attempts, and the number of points scored per position in the NBA. The results from this graph suggest that on average the Shooting guard has the highest percent 3 pt conversion out of all the positions which makes sense since their usual role in the team is to be able to score points from outside the box. Even so, it is interesting to see the position with the most attempts and lowest conversion rate from 3 pt is Guard as whereas I thought it was going to be the Center position. This is seen in the first graph since even tho the center position is the lowest line out of all the lines, the guard has a negative slope suggesting they are unreliable when it comes to 3 pt scored.

The results from the 2 pt attempts and the number of points scored per position were different which is expected. The most reliable position at scoring 2 pts is based on the graph the Guard position which was surprising after seeing the results of the last graph. The least reliable position at scoring 2 pts is the shooting guard which we can see in the graph since it has the lowest slope which was also really surprising but, makes sense since their usual role is to hit those 3-pointers.

```{r echo=FALSE}
library(GGally)
library(cluster)
library(ggplot2)
library(factoextra)
library(cluster)
library(NbClust)

#Created a linear regression + Scatterplot
ggplot(dfpergame,aes(x=PTS,y=X3PA,color=Pos))+theme_bw()+ggtitle("Number of points per position based on field goal attemps")+geom_point() + labs(x = "Three point attempts", y = "Points Scored", title = "Points scored vs 3pt attempt per position")+geom_smooth(method = "glm")

ggplot(dfpergame,aes(x=PTS,y=X2PA,color=Pos))+theme_bw()+ggtitle("Number of points per position based on field goal attemps")+geom_point() + labs(x = "Two point attempts", y = "Points Scored", title = "Points scored vs 2pt attempt per position") +geom_smooth(method = "glm")

```



Histogram+Scatterplot 

I wanted to see the relationship between 2-point attempts, 3-point attempts, and the amount of points players average. The results for this graph are interesting since the histogram suggests that players do attempt more 2 pts than 3 pts but the scatterplot suggests that players who attempt more 3pt do score more points at the end. 

```{r echo=FALSE}
# library
library(ggplot2)
library(ggExtra)
 
# The mtcars dataset is proposed in R
head(dfpergame)
 
# classic plot :
p <- ggplot(dfpergame, aes(x=X3PA, y=X2PA, color=PTS, size=X3PA)) +
      geom_point() +
      theme(legend.position="none")+ggtitle("Points convertion based on 2pt and 3pt attempts")
# Custom marginal plots:
p2 <- ggMarginal(p, type="histogram", fill = "slateblue", xparams = list(  bins=10))

p2

```

Histogram

The histograms show the number of points on average scored by each team in the 2020-2021 NBA season. The result shows that the team which scored the most points is Brooklyn Nets and the team that least scored points on average was TOT which in this dataset means players who played for multiple teams during the allotted season. So the true least scored point team during the allotted season was unfortunately Los Angeles Lakers (my team). These results surprised me since they did not do so bad compared to this new season since they finished 10th out of all the teams with the most amount of wins. It makes sense since most of that season one of its best players was injured for almost every single game.

```{r echo=FALSE}
library(ggplot2)


#Created ggplot histogram
ggplot(dfpergame,aes(x=Tm,y=PTS,color=Tm))+geom_bar(stat = "identity",width = .7)+theme_bw()+labs(title = "Amount points scored by Team")+xlab("Teams")+ylab("Points Scored")

```

\vspace{.2in}
\hrule
\vspace{.2in}

#  Conclusions 

This project was a lot of fun to do since it was open-ended as to we decide what we wanted to look for in a dataset that had to meet certain length requirements. The results from this project were eye-opening for me since I didn't expect most of them. Beginning with the average age by position, where the point guard has the most versatile age range and the players who play Forward and Center on average are the oldest were as I expected the Center position to be the oldest. This result made me realize that age is just a number and that players are either good at taking care of their body nowadays or that by changing their playstyle they can keep being in the NBA since the Point Guard position is one where players need to be either athletic or intelligent since they bring the ball up the court. 

Another eye-opening result came with the second and third graphs where I didn't expect to see the guard being the least efficient at 3 pt but most efficient at 2 pt range. Similarly, with the shooting guard, I did expect him to be the most efficient at 3 pt but not, the least efficient at 2 pt range. This tells me that for me to build the most efficient draft I have to be careful which player I have to pick for these positions since it is essential to the team. 

The histogram + scatterplot graph results were also unexpected in which it implies that the players who attempt more 3 pts end up scoring more points than the players who attempt more 2 pts on average. It makes sense since the game NBA has been moving towards a more 3pt game thanks to data analytics and it shows the efficiency through this cool graph. This finding made me realize that I need to find for my next draft players who attempt more 3pt than average to be able to have the best team possible just, not in the guard position seen in the results of earlier graphs.

The last graph was something I didn't expect. As a Lakers fan, it is hurtful to see that my team scored on average the least amount of points out of all the teams and that the Brooklyn Nets were the opposite. I will strongly consider this when drafting next season since, I do tend to bias Lakers players thinking they are the best, which by looking at the results is not the truth anymore.




\vspace{.2in}
\hrule
\vspace{.2in}



# Appendix

##Graph 1

library(ggplot2)

library(tidyverse)

library(dplyr)

library(olsrr)

library(pander)

#Reading the file

dfadplayers<-read.csv("/Users/eduardosalvador/Desktop/FINAL\ Spring\ Semester\ 2021/CMDA\ /Assignments/Project\ 1/archive/nba2021_advanced.csv",header = T)

dfpergame<-read.csv("/Users/eduardosalvador/Desktop/FINAL\ Spring\ Semester\ 2021/CMDA\ /Assignments/Project\ 1/archive/nba2021_per_game.csv",header = T)

ggplot(dfadplayers,aes(x=Pos,y=Age,fill=Pos))+theme_bw()+geom_boxplot()+ggtitle("Average player age based on position")

mean(dfadplayers$Age)

##Graph 2

#Created a linear regression + Scatterplot

ggplot(dfpergame,aes(x=PTS,y=X3PA,color=Pos))+theme_bw()+ggtitle("Number of points per position based on field goal attemps")+geom_point() + labs(x = "Three point attempts", y = "Points Scored", title = "Points scored vs 3pt attempt per position")+geom_smooth(method = "glm")

ggplot(dfpergame,aes(x=PTS,y=X2PA,color=Pos))+theme_bw()+ggtitle("Number of points per position based on field goal attemps")+geom_point() + labs(x = "Two point attempts", y = "Points Scored", title = "Points scored vs 2pt attempt per position") +geom_smooth(method = "glm")

##Graph 3

library(ggplot2)

library(ggExtra)
 
##Looking at the top 5 rows of data regarding the dfpergame

head(dfpergame)
 
##classic plot :

p <- ggplot(dfpergame, aes(x=PTS, y=X2PA, color=X3PA, size=X3PA)) +
      geom_point() +
      theme(legend.position="none")+ggtitle("Pts based on 2 pt and 3 pt attempts average per game")
      
##Custom marginal plots:

p2 <- ggMarginal(p, type="histogram", fill = "slateblue", xparams = list(  bins=10))

p2

##Graph 4

#Created ggplot histogram

ggplot(dfpergame,aes(x=Tm,y=PTS,color=Tm))+geom_bar(stat = "identity",width = .7)+theme_bw()+labs(title = "Amount points scored by Team")+xlab("Teams")+ylab("Points Scored")



\vspace{.2in}
\hrule
\vspace{.2in}




\clearpage




\clearpage



```



